---
title: "Inshore Marine Water Quality Exploratory Data Analysis"
subtitle: "A Healthy Waters Partnership Analysis"
description: "Script 4 in a series of script designed to analyse, score, and present inshore marine water quality in the Dry Tropics region. The output of this is used in the Dry Tropics Technical Report."
author: "Adam Shand"
format: html
params:
  project_crs: "EPSG:7844"
  target_fyear: 2024 
---

# Introduction

The purpose of this script is to create maps identifying the sampling site locations for the inshore marine water quality index.

These maps are used in the body of the methods document, and in the appendix of the technical report.

# Script Set Up

This script requires multiple core spatial packages, each of which is loaded below. Key variables are also created.

```{r}
#| label: load packages

#use pacman function to load and install (if required) all other packages
pacman::p_load(tidyverse, glue, here, sf, tmap, janitor, readxl, grid, RColorBrewer)

#load in the custom function used to create the read and write folders for the script
source(here("functions/script_setup.R"))

#run the function to create the folders and paths
script_setup()

#read in the custom function to clean column names into our specific style
source(here("functions/name_cleaning.R"))

#turn off spherical geometry
sf_use_s2(F)

#determine if we want the dataset to be the one with removed sites, or the original
removed <- params$sites_removed

#set the project crs
proj_crs <- params$project_crs

```

# Load Data

Now the script is set up we need to load in all of the required datasets. Spatial data for the northern three regions should be readily available in the repo, the dataset is created by the n3_region-builder.qmd script in the repo that should be the first script run for new users. If the dataset is not available refer back to the README document in the GitHub repo. The other parts of spatial data here should go off without a hitch if the region-builder script has been run.

```{r}
#| label: load in all data
#| output: false

#read in the northern three spatial files and reduce down to only the components we need
dt_marine <- n3_region |> 
  filter(Region == "Dry Tropics",
         Environment == "Marine", BasinOrZone != "Offshore") |> 
  mutate(WatercourseOrGeographicArea = case_when(WatercourseOrGeographicArea == "Open Coastal" ~ "H.Open Coastal",
                                     WatercourseOrGeographicArea == "Enclosed Coastal" ~ "H.Enclosed Coastal",
                                     T ~ WatercourseOrGeographicArea))

#get land components to help with background
dt_land <- n3_region |> 
  filter(Region == "Dry Tropics",
         Environment != "Marine") |> 
  group_by(BasinOrZone) |> 
  summarise(geom = st_union(geom)) |> 
  ungroup() |> st_cast() |> st_make_valid()

#read in site coords and convert to a simple feature object
sites <- read_excel(path = glue("{data_path}/dt_wq_inshore_metadata.xlsx"), sheet = "Current_Sites") |> 
  select(BasinOrZone, SubBasinOrSubZone, WatercourseOrGeographicArea, Code, Lat, Long) |> 
  st_as_sf(coords = c("Long", "Lat"), crs = proj_crs)

```

# Create Map

First we need to make sure that the colours for the background are correctly associated with each sub zone.

```{r}

#make columns factors
#sites <- sites |> 
  #mutate(WatercourseOrGeographicArea = as_factor(WatercourseOrGeographicArea))

#and for the background
#dt_marine <- dt_marine |> 
  #mutate(WatercourseOrGeographicArea = as_factor(WatercourseOrGeographicArea))




```

Now we can put everything together. We will loop over the Ross and Black to make individual maps.

```{r}
#| label: create region maps

#load in custom function to create the water related layers of the map
source(here("functions/maps_water_layer.R"))

#load in custom function to create the inset map
source(here("functions/maps_inset_layer.R"))

#create a background of the burdekin region for the inset map
dt_background <- n3_region |> 
  filter(Region == "Dry Tropics", Environment == "Marine") |> 
  group_by(Region) |> summarise(geom = st_union(geom))

#create a point of interest for townsville
tsv <- poi |> filter(PlaceName == "Townsville")

#create target vector to loop lover
target_vect <- unique(dt_marine$BasinOrZone)

#create a similar vect for land basins
land_vect <- rev(unique(dt_land$BasinOrZone))

colours_to_use <- list(c("#FBB4AE", "#B3CDE3", "#CCEBC5", "#DECBE4", "#FED9A6", "#FFFFCC"), #CB background
                       c("#FBB4AE", "#B3CDE3", "#CCEBC5", "#FED9A6", "#FFFFCC"), #CB sites
                       c("#B3CDE3", "#FFFFCC", "#DECBE4")) #HB background and sites

for (i in 1:length(target_vect)){
  
  #make a special colour counter that we can mess around with
  colour_counter <- i
  
  #filter sites and rename column
  filtered_sites <- sites |> filter(BasinOrZone == target_vect[i]) |> 
    rename("Geographic Area: Reported" = WatercourseOrGeographicArea)
  
  #define focus area for waterways and rename column
  focus_area <- dt_land |> filter(BasinOrZone == land_vect[i]) 
  
  #define focus area for marine wq sites and rename column
  focus_marine <- dt_marine |> filter(BasinOrZone == target_vect[i]) |> 
    rename("Geographic Area: All" = WatercourseOrGeographicArea)
  
  #create the base water layer for the specific basin
  maps_water_layer(stream_order = c(2,9), water_lines = T, water_polygons = T, 
                   basin_or_zone = land_vect[i], fast = T)
  
  #create the background layer using the broader dry tropics backdrop
  maps_inset_layer(supplied_sf_1 = focus_marine,
                   use_bbox_1 = FALSE,
                   supplied_sf_2 = filtered_sites,
                   use_bbox_2 = TRUE,
                   background = dt_background, aspect = 1)
  
  #adjust the colour call for the second loop
  if (colour_counter == 2) {colour_counter <- 3}
  
  #create the main map
  map <- tm_shape(qld) +
    tm_polygons(fill = "grey80", 
                col = "black") +
    tm_shape(filtered_sites, is.main = T) +
    tm_symbols(size = 0, 
               fill_alpha = 0) +
    tm_shape(dt_marine) +
    tm_polygons(fill = "white", 
                col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(fill = "grey90", 
                col = "grey90") +
    tm_shape(dt_land) +
    tm_polygons(fill = "grey90", 
                col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(fill = "Geographic Area: All", 
                fill.scale = tm_scale_categorical(values = colours_to_use[[colour_counter]]),
                fill_alpha = 0.7,
                col = "black") +
    water_map +
    tm_shape(tsv) +
    tm_symbols(size = 0.3, 
               fill = "white", 
               col = "black", 
               lwd = 1, 
               shape = 23) +
    tm_text("PlaceName", 
            size = 0.8,
            options = opt_tm_text(shadow = T, point.label = T))
  
  #adjust the colour call for the first loop
  if (colour_counter == 1) {colour_counter <- 2}
  
  map <- map  +
    tm_shape(filtered_sites) +
    tm_symbols(size = 0.6, 
               fill = "Geographic Area: Reported",
               fill.scale = tm_scale_categorical(values = colours_to_use[[colour_counter]]),
               col = "black", 
               lwd = 2, 
               shape = 21) +
    tm_text("Code", 
            size = 0.9,
            options = opt_tm_text(shadow = T, point.label = T)) +
    tm_layout(legend.bg.color = "white", 
              legend.position = c("right", "bottom"),
              asp = 1)
  
  #save map
  tmap_save(map, glue("{output_path}/{target_vect[i]}_Inshore.png"), insets_tm = inset_map, insets_vp = inset_viewport)
  
}

```

# Create De-identified Map

Below we will create the same maps, but now with sites de-identified.

```{r}
#| label: create deidentified region maps

#create target vector to loop lover
target_vect <- unique(dt_marine$BasinOrZone)

#create a similar vect for land basins
land_vect <- rev(unique(dt_land$BasinOrZone))

for (i in 1:length(target_vect)){
  
  #make a special colour counter that we can mess around with
  colour_counter <- i
  
  #filter sites and rename column
  filtered_sites <- sites |> filter(BasinOrZone == target_vect[i]) |> 
    rename("Geographic Area: Reported" = WatercourseOrGeographicArea)
  
  #define focus area for waterways and rename column
  focus_area <- dt_land |> filter(BasinOrZone == land_vect[i]) 
  
  #define focus area for marine wq sites and rename column
  focus_marine <- dt_marine |> filter(BasinOrZone == target_vect[i]) |> 
    rename("Geographic Area: All" = WatercourseOrGeographicArea)
  
  #create the base water layer for the specific basin
  maps_water_layer(stream_order = c(2,9), water_lines = T, water_polygons = T, 
                   basin_or_zone = land_vect[i], fast = T)
  
  #create the background layer using the broader dry tropics backdrop
  maps_inset_layer(supplied_sf_1 = focus_marine,
                   use_bbox_1 = FALSE,
                   supplied_sf_2 = filtered_sites,
                   use_bbox_2 = TRUE,
                   background = dt_background, aspect = 1)
  
  #adjust the colour call for the second loop
  if (colour_counter == 2) {colour_counter <- 3}
  
  #create the main map
  map <- tm_shape(qld) +
    tm_polygons(fill = "grey80", 
                col = "black") +
    tm_shape(filtered_sites, is.main = T) +
    tm_symbols(size = 0, 
               fill_alpha = 0) +
    tm_shape(dt_marine) +
    tm_polygons(fill = "white", 
                col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(fill = "grey90", 
                col = "grey90") +
    tm_shape(dt_land) +
    tm_polygons(fill = "grey90", 
                col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(fill = "Geographic Area: All", 
                fill.scale = tm_scale_categorical(values = colours_to_use[[colour_counter]]),
                fill_alpha = 0.7,
                col = "black") +
    water_map +
    tm_shape(tsv) +
    tm_symbols(size = 0.3, 
               fill = "white", 
               col = "black", 
               lwd = 1, 
               shape = 23) +
    tm_text("PlaceName", 
            size = 0.8,
            options = opt_tm_text(shadow = T, point.label = T))
  
  #adjust the colour call for the first loop
  if (colour_counter == 1) {colour_counter <- 2}
  
  map <- map  +
    tm_shape(filtered_sites) +
    tm_symbols(size = 0.6, 
               fill = "Geographic Area: Reported",
               fill.scale = tm_scale_categorical(values = colours_to_use[[colour_counter]]),
               col = "black", 
               lwd = 2, 
               shape = 21) +
    #tm_text("Code", 
    #        size = 0.9,
    #        options = opt_tm_text(shadow = T, point.label = T)) +
    tm_layout(legend.bg.color = "white", 
              legend.position = c("right", "bottom"),
              asp = 1)
  
  #save map
  tmap_save(map, glue("{output_path}/{target_vect[i]}_Inshore_Deidentified.png"), insets_tm = inset_map, insets_vp = inset_viewport)
  
}

```

