---
title: "Pesticides Data Analysis"
subtitle: "A Healthy Waters Partnership Analysis"
description: "This script analyses the pesticide sampling locations in the Dry Tropics region. The output of this is used in the Dry Tropics Technical Report."
author: "Adam Shand"
format: html
params:
  project_crs: "EPSG:7844"
  target_fyear: 2024 
---

# Introduction

The purpose of this script is to analyse the pesticides data from the sampling site locations for the freshwater pesticides index and calculate standardised scores for the Dry Tropics Technical Report. This data is used in the body of the technical report as well as the appendix.

# Script Set Up

This script requires multiple core spatial packages, each of which is loaded below. Key variables and paths are also created.

```{r}
#| label: load packages

#use pacman function to load and install (if required) all other packages
pacman::p_load(tidyverse, glue, here, janitor, readxl)

#load in the custom function used to create the read and write folders for the script
source(here("functions/script_setup.R"))

#run the function to create the folders and paths
script_setup()

##read in the custom function to clean column names into our specific style
source(here("functions/name_cleaning.R"))

#set project variables
proj_crs <- params$project_crs

```

# Load Data

Now the script is set up we need to load in all of the required datasets. For this script, we will only need the metadata and master data spreadsheets.

```{r}
#| label: load in all data

#read in the target sheet and drop unwanted column names, then run the custom naming cleaning function
site_metadata <- read_excel(glue("{data_path}/dt_wq_freshwater_pesticides_metadata.xlsx"),
                            sheet = "Current_Sites", na = c("", "NA", "NULL", "null")) |> 
  select(-c(SiteName, AKA)) |> 
  name_cleaning()


#read in the target sheet and drop unwanted column names, then run the custom naming cleaning function
site_data <- read_excel(glue("{data_path}/dt_wq_freshwater_pesticides_data_master.xlsx"), 
                         sheet = "CLMP", na = c("", "NA", "NULL", "null")) |> 
  select(-SiteName) |> 
  name_cleaning()
  
#join metadata and results together
pesticides_all <- site_metadata |> 
  left_join(site_data)

```

# Analyse Data

Given the simplicity of the dataset we can move straight to calculating the final pesticides score. This is achieved using a custom scoring function that we create below.

```{r}
#| label: analyse data

#create a standardizing function
standardise_scores <- function(col){
  col = case_when(col <= 1 ~ 81 + abs((19 - ((col - 0) * (19/1)))),
                  col > 1 & col < 5 ~ 61 + abs((19.9 - ((col - 1.01) * (19.9/3.99)))),
                  col >= 5 & col < 10 ~ 41 + abs((19.9 - ((col - 5) * (19.9/4.99)))),
                  col >= 10 & col < 20 ~ 21 + abs((19.9 - ((col - 10) * (19.9/9.99)))),
                  TRUE ~ 0 + abs((20.9 - ((col - 20) * (20.9/79.99)))))
}

#run the function
pesticides_all <- pesticides_all |> 
  mutate(Score = standardise_scores(MsPaf))

#assign a grade
pesticides_all <- pesticides_all |> 
  mutate(Grade = case_when(Score >= 81 ~ "A",
                           Score >= 61 ~ "B",
                           Score >= 41 ~ "C",
                           Score >= 21 ~ "D",
                           T ~ "E"))

```

# Save Data

And now we can save the results, we will use a custom function to colour the cells appropriately.

```{r}
#| label: save results

#save the entire dataset
write_csv(pesticides_all, glue("{output_path}/pesticides_full_results.csv"))

#cut down dataset to only the proportions
pesticides_percent_only <- pesticides_all |> 
  filter(str_detect(Date, as.character(current_fyear))) |> 
  select(BasinOrZone, Date, ProportionInsecticides, ProportionOtherHerbicides, ProportionPsiiHerbicides, SpeciesProtected)

#and save that as well
write_csv(pesticides_percent_only, glue("{output_path}/pesticides_percentages.csv"))

#cut down dataset to the essentials for the report
pesticides_scores_only <- pesticides_all |> 
  filter(str_detect(Date, as.character(current_fyear))) |> 
  select(BasinOrZone, Date, Score, Grade, Risk)

#load in custom function
source(here("functions/cond_form_rc_grades.R"))

#save data
cond_form_rc_grades(pesticides_scores_only, glue("{output_path}/pesticide_scores"), cols = 3, method = "Numeric")

```

